---
title: "EDA for NYCDSA Shiny Project"
author: "Ryan Willett"
output: html_notebook
---

```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(dendextend)
library(ape)
```


Read in expression data from the data file. It is currently arranged to that the yeast strains (observations) are in the column and the gene names (variables) are in the row. I applied a transpose to switch the columns and rows.
```{r}
data <- read.csv(file="./data/SC_expression.csv", stringsAsFactors = F, row.names = 1, header = T)
conditions <- read.csv(file="./data/conditions_annotation.csv", stringsAsFactors = F, header = T)
data <- as.data.frame(t(data))
```


```{r}
# Inspecting that the data looks OK
data[1:20, 1:10]
```


Bringing in the condition data labels
```{r}
data <- data %>% rownames_to_column("ID") # Making ID column from the rownames so I can join with conditions table

data <- left_join(data, conditions, by="ID") # Binding the experimental conditions to the expression data

data$additional_information <- NULL # Don't need this column and it is mostly blank
```

Cleaning up the categories
```{r}
data <- data %>% 
  mutate(primary = gsub("wild type", "wildtype", primary)) %>% # Aligning the format of control name
  mutate(primary = gsub("^YB.*", "Biofuel.Production.Strain", primary)) %>%
  mutate(primary = gsub("^37.*", "37 deg", primary)) %>% # These were not grouped properly. Renamed to align
  mutate(primary = gsub("^Strain.*", "Phenol.Lysis", primary)) %>% 
  mutate(primary = gsub("^DDY.*", "Strain.W303", primary)) %>% 
  mutate(primary = gsub("^E.*", "Ethanol", primary)) %>% 
  mutate(primary = gsub("^G.*", "Glucose", primary)) %>% 
  mutate(primary = gsub("^hybrid.*", "hybrid cerevisiae x paradoxus", primary)) %>% select(-secondary, -ID) %>% 
  filter(!is.na(primary))


```


```{r}
# To explore the categories in the dataset
data %>% group_by(primary) %>% tally() 

levels(factor(data$primary))
```

Filtering with the inputs from the user
```{r}
input1 <- "wildtype"
input2 <- "Biofuel.Production.Strain"

data_subcat <- data %>% filter(primary == input1 | primary == input2) # Filtering for the groups to be compared
```

Evaluation of the statistical difference between the selected groups. This will be between 2 groups, so it's a t-test
```{r}
p_eval <- function(x, y) {
  ttest_res <- t.test(x,y) # t-test against the 2 categories to determine significance
  p.val <- ttest_res$p.val # Extracting the p-value from the t.test output
  return(p.val)
}
```


```{r}
#Descriptive statistics

data_means <- data %>%
  group_by(primary) %>% 
  summarise_all(list(mean = mean)) %>% 
  column_to_rownames("primary") %>% 
  t() %>% as.data.frame()  # Transposing the df (converts to matrix) and reclassifying as a df

data_sds <- data %>% 
  group_by(primary) %>% 
  summarise_all(list(sd = sd)) %>% 
  column_to_rownames("primary") %>% 
  t() %>% as.data.frame()  # Transposing the df (converts to matrix) and reclassifying as a df

```

```{r}

subcat_in1 <- data_subcat %>% filter(primary == input1)

subcat_in2 <- data_subcat %>% filter(primary == input2)

pstats <- c()

for (c in 1:(ncol(data_subcat)-1)){ 
  test1 <- subcat_in1[,c]
  test2 <- subcat_in2[,c]
  
  p <- p_eval(test1, test2)
  pstats[c] <- p # Build up p-values list into a vector
}
```

```{r}
data_means_sub <- data_means %>% select(input1, input2) %>% rownames_to_column("genes") %>% mutate(genes = gsub("_mean", "", genes))

data_means_sub <- cbind(data_means_sub, pstats)

data_means_sub <- data_means_sub %>% 
  mutate(fold.change = data_means_sub[,3]/data_means_sub[,2],
         l2fc = (log2(data_means_sub[,3]) - log2(data_means_sub[,2])), 
         neg_pL10 = -1*log10(pstats))



```

```{r}
t <- theme_classic()
plot_colors <- scale_color_manual(values = c("black", "red", "green"))
#### Creating test p-values for plotting ####

data_means_sub <- data_means_sub %>% mutate(col_flag = ifelse(((l2fc>=1 | l2fc<=-1) & neg_pL10 > 1.3),1,0))

sig_genes <- data_means_sub %>% filter(col_flag == 1) %>% select(-col_flag)

ggplot(data_means_sub, aes(x=l2fc, y=neg_pL10, col=as.factor(col_flag))) + t + 
  geom_point(size=0.2) + #+scale_y_continuous(trans="log10")
  plot_colors +
  labs(x = "log2 Fold Change in Gene Expression", y = "-log10(p-value)")
```

Clustering analysis
```{r}
data_raw <- data %>% select(-primary) # Removing categorical data so they are not used in the clustering

d <- dist(as.matrix(data_raw)) #cancer_data is used so as not to bias the clustering with the known tumor classification in cancer_join
cat_n <- 4
hc <- hclust(d, method="ward.D")
hc <- as.dendrogram(hc)

#Plot the data
#cols_branches <- c("blue", "green", "red", "black", "orange")
dend <- color_branches(hc, k=cat_n, col = c("blue", "green", "red", "black", "orange"))
#labels(dend) <- NULL # Remove labels because they make the image too busy
plot(dend, main = "Clustering Based on 4 Classes")


# circlize_dendrogram(dend)
plot(dend, type = "triangle", ylab = "Height")

dend <- dend %>% set("labels", "")
rad_dend_col <- c("blue", "green", "red", "black", "orange")
clus <- cutree(hc, 4)
plot(as.phylo(hc), type="fan", tip.color = rad_dend_col[clus], label.offset = 1, main = "Clustering Based on 4 Classes")

ggplot(dend) +
  coord_polar(theta = "x") +
  scale_y_reverse(expand = c(0.2, 0))
```



